One-dimensional Classical Gas in a Harmonic Trap with Repulsive Interaction

### 2017-12-05 draft_05.tex
working on the breathing mode section

In the parameter regime where the energy $E$ is very large ($E\gg F_0$), it is dominated by the trap
energy, i.e., the interaction energy is negligible.  Thus,
$E\approx\omega_0^2\sum_{i}x_i^2=N\omega_0^2R^2$ in this regime.

@@

%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
\begin{figure*}
\includegraphics[width=0.9\textwidth]{ZhiyuPictures/rotating_frame_t=0-80T.eps}  
\caption{Phase space snapshots of the cloud dynamics viewed in the rotating frame, at longer
  timescales.}
\label{fig:Breathingfrequency2_1}
\end{figure*}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%

@@

%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
\begin{figure}[h]
\center
\includegraphics[scale=0.32]{ZhiyuPictures/freq_scanF_differentE_log_2_with_oscillation_demo.png}
\caption{Breathing mode frequency measured at different $E$ and $F_0$. Inset: a demonstration of the
  oscillation of $E_{trap}$, the total potential energy of particles in the trap, Since
  $E_{trap}\propto \sum_{is}{x_i^2}$, it manifests the oscillation behavior of the radius $R(t)$ as
  defined in Eq.~\ref{eq:def_of_R}.  The breathing frequency is obtained from the Fourier transform
  of $R(t)$.}
\label{fig:Breathingfrequency1}
\end{figure}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%

@@

Without interaction, the breathing mode frequency is exactly 2.  This is visualized readily from the
top panel of \ref{fig:Breathingfrequency2_0}, where $R(t)$ is the extent of the distribution in the
horizontal ($X$) direction.  As the line rotates clockwise with frequency $\omega=1$, within each
period the line is twice horizontally aligned (maximum $R$) and twice vertically aligned (minimum
$R$), so that the frequency of $R(t)$ is $\omega_0=2$.  When there is interaction, the frequency
will get shifted: $\omega_0 = 2+\delta$.  We measure the radius of the cloud $R(t)$ and get the
frequency spectrum of its oscillation behavior by Fourier transform. Then we take the peak frequency
near 2 as the breathing mode frequency.

@@

@@

To see this picture better, let us first introduce a approach to simplify this problem:

\subsection{Rotating frame in Phase-space}
If there is a non-circular distribution cloud in phase-space, e.g., a line-shape cloud in
Fig.\ref{fig:Breathingfrequency2_0}, driven by the harmonic trap, the cloud will rotate at frequency
$\omega_0$. When there is interaction, we may think there is a small modification of the frequency
$\omega=\omega_0+\delta$. Now we want to understand how $\delta$ is dependent on parameters, so it
is better to stand in the rotating frame in the phase-space. In this frame, the picture we will see
is as follow:\\
\begin{enumerate}[\textbf{*}]
\item All particles are stationary when there is no interaction.
\item The real X and P axis are rotating counter-clockwise, using real X axis to measure distance between particles to determine the interaction at each moment.
\item When there is interaction, particles will gain a ``velocity" in phase space: $\dot{P}=F/m$ . Other effects are all cancelled by frame rotation.
\end{enumerate}

In our system, we choose initial velocity to be zero, which means, the particles are aligned on $X|_{t=0}$ axis in phase-space at the beginning. If the breathing frequency of the system is $\omega=\omega_0+\delta$, we will expect to see a line (which may gradually deform into an oval cloud or even an isotropic cloud) rotating at $\delta$ in rotating frame. Since the motion we see is an additional rotation in rotating frame, we will call it precession. This precession directly leads to the $\delta$ shift in breathing frequency. Fig.\ref{fig:Breathingfrequency2_1} show this precession motion in the rotating frame in phase-space.

\subsection{Estimating Breathing Frequency}
Let us begin with two-particle case. When interaction is very strong, the effect of interaction is equivalent to exchanging momentum when two particles approach each other at distance $\sigma$. In the rotating frame in phase-space, this process can be interpret as follows:

\begin{figure}[h
]
\subfigure[]{
\begin{minipage}[b]{0.4\textwidth}
\includegraphics[width=1\textwidth]{ZhiyuPictures/bounce.png}
\end{minipage}
}
\subfigure[]{
\begin{minipage}[b]{0.4\textwidth}
\includegraphics[width=1\textwidth]{ZhiyuPictures/pass.png}
\end{minipage}
}
\caption{a schematic diagram of the collision process: \textbf{(a)}When $F_0\sigma$ is large, particles can not pass each other. Subduing to a constant repulsive force in collision, their trajectory in rotating frame is a circular arc.  \textbf{(b)} When $F_0\sigma$ is small, particles easily pass each other, so that the force they feel is not continuous. Their trajectory has a turning point. The dashed bright blue trajectory is identical with the one in (a), which is the critical situation between passing and bouncing. In that case, particles have zero relative velocity when they collide. }
\label{fig:Breathingfrequency3}
\end{figure}

(See Fig.\ref{fig:Breathingfrequency3}(a))Two particles flip to another line(thick gray line) which deviate a small angle $\alpha$ away from original configuration. Similar process happen when there are more particles. In every period of harmonic oscillation, each particle meet N particles and collide 2N times, while half of the collisions (N times) are between particles with huge difference in momentum, which is just the process shown in Fig.\ref{fig:Breathingfrequency3}. (As for the remaining half of collisions, colliding particles have small difference in their momentum. Let us say, both particles have positive momentum. In Fig.\ref{fig:Breathingfrequency3}, it can be shown that two particles on the same half of the P-axis seldom contribute to the rotation of distribution.)

Let us estimate the precession angle $\alpha$ in one collision event by $\frac{\sigma}{R}$, which follows from the two-particle case study. Radius of the cloud R can be estimated according to $E=Nm\omega_0^2R^2$
We will get the precession angular velocity 
\begin{equation}
\delta=2\omega_0\frac{N^\frac{3}{2}m^{\frac{1}{2}}E^{-\frac{1}{2}}\sigma\omega_0}{2\pi}
\label{eq:breathingfrequency1}
\end{equation}

>>>>

In the dilute gas limit, there are few or no interactions occuring at most times.  Thus the energy
$E$ is dominated by the trap energy, i.e., $E\approx m\omega_0^2\sum_{i}x_i^2= Nm\omega_0^2R^2$, or,
in our units, $E\approx NR^2$.

@@

%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
\begin{figure*}
\includegraphics[width=0.9\textwidth]{ZhiyuPictures/rotating_frame_t=0-80T.eps}  
\caption{Phase space snapshots of the cloud dynamics viewed in the rotating frame, at longer
  timescales.  The broadening and rotation observed here are interaction effects: in the
  non-interacting gas the particles are stationary in this frame.}
\label{fig:Breathingfrequency2_1}
\end{figure*}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%

@@

%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
\begin{figure}[h]
\center
\includegraphics[scale=0.32]{ZhiyuPictures/freq_scanF_differentE_log_2_with_oscillation_demo.png}
\caption{Breathing mode frequency measured at different $E$ and $F_0$, for $N=5$ particles. Inset: a
  demonstration of the oscillation of $E_{trap}$, the total potential energy of particles in the
  trap, Since $E_{trap}\propto \sum_{is}{x_i^2}$, it manifests the oscillation behavior of the
  radius $R(t)$ as defined in Eq.~\ref{eq:def_of_R}.  The breathing frequency is obtained from the
  Fourier transform of $R(t)$.}
\label{fig:Breathingfrequency1}
\end{figure}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%

@@

Without interaction, the breathing mode frequency is exactly 2.  This is visualized readily from the
top panel of \ref{fig:Breathingfrequency2_0}, where $R(t)$ is the extent of the distribution in the
horizontal ($X$) direction.  As the line rotates clockwise with frequency $\omega=1$, within each
period the line is twice horizontally aligned (maximum $R$) and twice vertically aligned (minimum
$R$), so that the frequency of $R(t)$ is $\omega_B=2$.  When there is interaction, the frequency
will get shifted: $\omega_B = 2+\delta$.  We measure the radius of the cloud $R(t)$ and get the
frequency spectrum of its oscillation behavior by Fourier transform. Then we take the peak frequency
near 2 as the breathing mode frequency.

@@

\subsection{Phase space estimate}

Since we are interested in the deviation $\delta=\omega_B-2$ from the non-interacting breathing
frequency, it is useful to work in the rotating frame, in which a non-interacting cloud would be
stationary (non-rotating).  The rotation frequency of the cloud in this frame is then $\delta$.

%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
\begin{figure}[h
]
\subfigure[]{
\begin{minipage}[b]{0.4\textwidth}
\includegraphics[width=1\textwidth]{ZhiyuPictures/bounce.png}
\end{minipage}
}
\subfigure[]{
\begin{minipage}[b]{0.4\textwidth}
\includegraphics[width=1\textwidth]{ZhiyuPictures/pass.png}
\end{minipage}
}
\caption{a schematic diagram of the collision process: \textbf{(a)}When $F_0\sigma$ is large, particles can not pass each other. Subduing to a constant repulsive force in collision, their trajectory in rotating frame is a circular arc.  \textbf{(b)} When $F_0\sigma$ is small, particles easily pass each other, so that the force they feel is not continuous. Their trajectory has a turning point. The dashed bright blue trajectory is identical with the one in (a), which is the critical situation between passing and bouncing. In that case, particles have zero relative velocity when they collide. }
\label{fig:Breathingfrequency3}
\end{figure}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%

To analyze the rotation relative to the $X_{\rm rot}$-$P_{\rm rot}$ frame, we consider two-particle
collisions.  In Figure \ref{fig:Breathingfrequency3}(a), we show schematics of such a collision near
the trap center at strong interactions.  For very large $F_0$, the particles simply change momentum
when they are at distance $\sigma$ from each other; this process is shown in red as two straight
lines.  (The particles are initially at $X=\pm\frac{1}{2}\sigma$ and remain at these $X$ values,
but exchange their momenta.)   For smaller $F_0$, there is change of both position and momentum, as shown in yellow, green
and blue for successively weaker interactions. The blue line shows a case where $F_0$ is smaller
enough such that the particles can actually cross each other.

In Figure \ref{fig:Breathingfrequency3}(a) we show collisions for small $F_0$, so that the particles
pass each other.  in the $X_{\rm rot}$-$P_{\rm rot}$ plane, this is seen as a turning point in the trajectory.

We now estimate $\delta$ at large $F_0$.  The relevant collision process is that shown by the red
lines in Figure \ref{fig:Breathingfrequency3}(a).  The two particles, which are initially on the
$X_{\rm rot}$ axis, are moved to two other points, on the thick gray line, which deviates a small
angle $\alpha$ away from original configuration.  In every period of the harmonic oscillation, each
particle meets each of the other particles twice.  Half of these collisions (about $N$ collision)
are between particles with large difference in momentum, which is the process shown in
Fig.\ref{fig:Breathingfrequency3}.  (As for the remaining half of the collisions, colliding
particles have smaller difference in their momentume, e.g., both particles having positive
momentum.  In this estimate we  ignore the effects of these collisions, as they clearly contribute
far less to the rotation of the particle distribution.)

The precession angle is $\alpha\sim \sigma/P$.  In our rescaled units, the momenta that the
particles have near $x=0$ are of the order $R$, the size of the cloud.  (In phase space $R$ can be
interpreted either as the extent in real space when the velocities are small, or as the extent in
momentum space when the particles approach $x=0$, i.e. when the distribution is along the $P$
direction.)  Thus, we can estimate  $\alpha\sim \sigma/R$, i.e.,   $\alpha\sim 1/R$, since our unit
of length is $\sigma$.  This is the rotation per unit collision.  Since there are $\sim N$
collistions per period $T=2\pi$, we have the estimate for the interaction-induced rotation per unit
time as
\begin{equation}
\delta ~=~ 2 \frac{N}{2\pi} \frac{1}{R} ~=~  \frac{1}{\pi}  N^{3/2}E^{-1/2}
\label{eq:breathingfrequency1}
\end{equation}
The factor $2$ accounts for the fact that each rotation of the elongated cloud in phase space
corresponds to two breathing mode periods.  In the last step we have used the estimate $E\sim
NR^2$.

@@

%% \subsection{Estimating Breathing Frequency}

%% Let us begin with two-particle case. When interaction is very strong, the effect of interaction is
%% equivalent to exchanging momentum when two particles approach each other at distance $\sigma$. In
%% the rotating frame in phase-space, this process can be interpret as follows:

%% (See Fig.\ref{fig:Breathingfrequency3}(a))Two particles flip to another line(thick gray line) which
%% deviate a small angle $\alpha$ away from original configuration. Similar process happen when there
%% are more particles. In every period of harmonic oscillation, each particle meet N particles and
%% collide 2N times, while half of the collisions (N times) are between particles with huge difference
%% in momentum, which is just the process shown in Fig.\ref{fig:Breathingfrequency3}. (As for the
%% remaining half of collisions, colliding particles have small difference in their momentum. Let us
%% say, both particles have positive momentum. In Fig.\ref{fig:Breathingfrequency3}, it can be shown
%% that two particles on the same half of the P-axis seldom contribute to the rotation of
%% distribution.)

%% Let us estimate the precession angle $\alpha$ in one collision event by $\frac{\sigma}{R}$, which
%% follows from the two-particle case study. Radius of the cloud R can be estimated according to
%% $E=Nm\omega_0^2R^2$ We will get the precession angular velocity
%% \begin{equation}
%% \delta=2\omega_0\frac{N^\frac{3}{2}m^{\frac{1}{2}}E^{-\frac{1}{2}}\sigma\omega_0}{2\pi}
%% \label{eq:breathingfrequency1}
%% \end{equation}
